---
title: "p8105_hw2_tl3196"
author: "Tianshu Liu"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, collapse = FALSE, warning = FALSE)
options(tibble.print_min = 10)
```

## Problem 1

```{r subway_station}
library(tidyverse)
sub_stat_df = 
  read_csv('data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv',
  col_types = cols(Route8 = "c", Route9 = "c", Route10 = "c", Route11 = "c")) %>%
  janitor::clean_names() %>%
  mutate(
    entry = recode(entry, 'YES' = TRUE, 'NO' = FALSE)
  ) %>%
  select(line:entry, vending, ada)

sub_stat_df
```

There are `r nrow(sub_stat_df)` × `r ncol(sub_stat_df)` data in the dataset.

The variables contains `r colnames(sub_stat_df)`.

This dataset is not tidy.

```{r cal_dist, include=FALSE}
# calculate distinct stations
num_stat = nrow(
  sub_stat_df %>% 
    distinct(station_name, line)
)

# calculate distinct stations with ADA
num_ada = nrow(
  sub_stat_df %>%
    distinct(line, station_name, ada) %>% 
    filter(ada == TRUE)
)

# calculate proportion of entrances/exits without vending allow entrance
prop_no_vend = nrow(sub_stat_df %>%
  filter(vending == "NO", entry == "TRUE")
) / nrow(sub_stat_df %>%
  filter(vending == "NO")
)
```

There are `r num_stat` subway stations in NY, `r num_ada` of which are ADA compliant.

`r prop_no_vend*100` % of the station entrances / exits without vending allow entrance.

```{r clean_sub_data}
# clean data with pivot_longer function
clean_sub_stat_df = sub_stat_df %>% 
  pivot_longer(route1:route11,
    names_to = "route_num",
    values_to = "route",
    values_drop_na = TRUE)

# distinct stations serve A train
clean_sub_stat_df %>%
  filter(route == "A") %>% 
  distinct(station_name, line)

# distinct stations serve A train with ADA
clean_sub_stat_df %>%
   filter(route == "A", ada == "TRUE") %>% 
  distinct(station_name, line)
```

## Problem 2
Import datasets from `Trash Wheel Collection Data.xlsx` file, and clean them.

```{r import&clean_trash_data}
library(readxl)
mr_trash_whl = read_excel(
  "./data/Trash Wheel Collection Data.xlsx",
  sheet = 1,
  range = cell_cols("A:N"),
  skip = 1,
  na = ""
  ) %>%
  janitor::clean_names() %>%
  drop_na(dumpster) %>%
  mutate(
    sports_balls = as.integer(sports_balls),
    year = as.numeric(year)
  )

mr_trash_whl

prof_trash_whl = read_excel(
  "./data/Trash Wheel Collection Data.xlsx",
  sheet = 2,
  range = cell_cols("A:M"),
  skip = 1,
  na = ""
  ) %>%
  janitor::clean_names() %>%
  drop_na(dumpster) %>%
  mutate(
    sports_balls = NA,
    year = as.numeric(year)
  )

prof_trash_whl
```

Add a column `trash_wheel` to keep track of the dumpster comes from which trash wheel, and full_join two datasets to the new dataset `trash_whl`.

```{r combine_df}
trash_whl = 
  full_join(
    mr_trash_whl %>% 
      mutate(
        trash_wheel = "mr_trash_wheel"
        ), 
    prof_trash_whl %>% 
      mutate(
        trash_wheel = "prof_trash_wheel"
        )
    ) %>%
  relocate(trash_wheel)

trash_whl

# calculate total weight of trash collected by Mr. Trash Wheel
mr_trash_weight = sum(
  trash_whl %>%
  filter(trash_wheel == 'mr_trash_wheel') %>%
  select(weight_tons)
  )

# calculate total weight of trash collected by Professor Trash Wheel
prof_trash_weight = sum(
  trash_whl %>%
  filter(trash_wheel == 'prof_trash_wheel') %>%
  select(weight_tons)
  )

# calculate total number of sports balls collected by Mr. Trash Wheel in 2020
mr_sports_ball_2020 = sum(
  trash_whl %>%
  filter(trash_wheel == 'mr_trash_wheel', year == 2020) %>%
  select(sports_balls)
)
```
This data set displays the trash record collected by `Mr. Trash Wheel` and `Professor Trash Wheel` in every dumpster.
The combined data set is ``r nrow(trash_whl)``×``r ncol(trash_whl)``.

Each record in the combined data set represents a dumpster collected by a trash wheel.
Among ``r nrow(trash_whl)`` dumpsters, ``r nrow(filter(trash_whl, trash_wheel == 'mr_trash_wheel'))`` dumpsters come from Mr. Trash Wheel, and ``r nrow(filter(trash_whl, trash_wheel == 'prof_trash_wheel'))`` dumpsters come from Professor Trash Wheel.

The variables in the combined data set are ``r colnames(trash_whl)``.
Variable `trash_wheel` indicates which trash wheel the record comes from.
Variable `month, year, date` introduces the date of a dumpster.
The rest variables introduce the statistical information of each dumpster.

Mr and Professor Trash Wheel have collected ``r sum(select(trash_whl, weight_tons))`` tons of trash in total. Among them, Mr. Trash Wheel has collected ``r mr_trash_weight`` tons of trash, and Professor Trash Wheel has collected ``r prof_trash_weight`` tons of trash.

Mr. Trash Wheel has totally collected ``r mr_sports_ball_2020`` sports balls in `2020`.

## Problem 3
First, clean the data in pols-month.csv.

```{r clean_pols_month_data}
# First, clean the data in pols-month.csv.
pols_df = read_csv("./data/fivethirtyeight_datasets/pols-month.csv") %>%
  janitor::clean_names() %>%
  separate(col = mon, into = c("year", "month", "day"), sep = "-", convert = TRUE) %>%
  arrange(year, month) %>%
  mutate(
    month = month.name[.$month],
    president = ifelse(prez_gop, "gop", "dem")
  )%>%
    select(-day, -prez_gop, -prez_dem) %>%
  relocate(president, .after = month)

pols_df
```

Second, clean the data in snp.csv. 

```{r clean_snp_data}
# Second, clean the data in snp.csv. 
snp_df = read_csv("./data/fivethirtyeight_datasets/snp.csv")  %>%
  janitor::clean_names() %>%
  separate(date, into = c("month", "day","year"), sep = "/", convert = TRUE) %>%
  mutate(year = ifelse(year>22, year + 1900, year + 2000)) %>%
  arrange(year, month) %>%
  mutate(month = month.name[.$month]) %>%
  select(-day) %>%
  relocate(year, month)

snp_df
```

Third, clean the data in unemployment.csv.

```{r clean_unemployment_data}
# Third, clean the data in unemployment.csv.
unemploy_df = 
  read_csv(
  "./data/fivethirtyeight_datasets/unemployment.csv",
  skip = 1,
  col_names = c("year", 1:12)) %>%
  janitor::clean_names() %>%
  pivot_longer(
    "x1":"x12", 
    names_to = "month", 
    values_to = "perc_of_unem", 
    names_prefix = "x",
    names_transform = list(month = as.numeric)) %>%
  arrange(year, month) %>%
  mutate(month = month.name[.$month]) 

unemploy_df
```

Left join 3 dfs.

```{r left_join_dfs}
# Left join 3 dfs
left_join_df = pols_df %>%
  left_join(snp_df) %>%
  left_join(unemploy_df)

left_join_df
```

The pols dataset contains ``r nrow(pols_df)`` records from ``r min(pols_df$year)`` to ``r max(pols_df$year)``.
The data includes ``r ncol(pols_df)`` variables, they are ``r colnames(pols_df)``. 

The snp dataset contains ``r nrow(snp_df)`` records from ``r min(snp_df$year)`` to ``r max(snp_df$year)``.
The data includes ``r ncol(snp_df)`` variables, they are ``r colnames(snp_df)``. 

The unemployment dataset contains ``r nrow(unemploy_df)`` records from ``r min(unemploy_df$year)`` to ``r max(unemploy_df$year)``.
The data includes ``r ncol(unemploy_df)`` variables, they are ``r colnames(unemploy_df)``. 

After left-joining, the joined dataset contains ``r nrow(left_join_df)`` records from ``r min(left_join_df$year)`` to ``r max(left_join_df$year)``.
The dataset includes ``r ncol(left_join_df)`` variables, they are ``r colnames(left_join_df)``.

The function `left_join` effectively helps merge datasets and maintain all the record of dates in the first dataset which is `pol_df`. 
However, for the dates that do not exist in the last two datasets, the data in the column `close, perc_of_unem` remains `NA`, which is not good for subsequent use.
Thus, try `inner_join` to join these three datasets.

```{r inner_join_dfs}
# Inner join 3 dfs
inner_join_df = pols_df %>%
  inner_join(snp_df) %>%
  inner_join(unemploy_df)

inner_join_df
```

After inner_joining, the joined dataset contains ``r nrow(inner_join_df)`` records from ``r min(inner_join_df$year)`` to ``r max(inner_join_df$year)``, which only contains records whose date exists in all three datasets.
The dataset still includes ``r ncol(inner_join_df)`` variables, they are ``r colnames(inner_join_df)``. 
